# Import, visualize and explore data

```{block, type='rmdoutcomes'}
- Import data into R
- Explain the general call to the function `ggplot()`
- Plot 5 frequently used types of graphics
```

## Data import from text files

To import a data set from a text file (e.g. csv, .txt, .dat) into R, we will use the library `readr` which is part of the `tydiverse`. We first load the library.

```{r, message = F, warning = F}
library(tidyverse)
```

Let's assume that the data is stored in the folder `data`. If this is not your case, change the path accordingly. To load the data, we can choose among several functions that all start with `read_`. The most generic one is `read_delim()` where we can specify how the columns are separated (delimited) in the data file.


```{r}
emissions <- read_delim(file = 'data/emissions.csv', delim = ';')
```

The path above looks for a folder called `data` in your working directory. The working directory is the local folder on your computer that that R will use for reading and writing files. Working with working directories is very ueful, because you don't have the write the full path to a data file, for example `C:/Users/...`. These file directories will be different for each user and computer. The code chunk above here uses a relative directory, which starts in the working directory. This makes sharing code very easy, as long as everyone is using the same folder structure, because you don't have to specify the folder where the data is stored. 

If the folder `data` is not stored in your current working directory, but in a folder one level up, you should add two dots to the directory above: `emissions <- read_delim(file = '../data/emissions.csv', delim = ';')`. This will R to search for the folder named `data` at another location. 


Let's have a look at the data. This is a data set on greenhouse gas emissions by source sector in the EU that I downloaded from [eurostat](https://appsso.eurostat.ec.europa.eu/nui/show.do?dataset=env_air_gge&lang=en) on 2021-04-30. It contains the greenhouse gas emissions in CO~2~ equivalent, in Mio tonnes, per vehicle type. The database could be a great source for data in your reports `r emo::ji('smile')`.

```{r}
emissions
```

The result of reading the data with any function in the library `readr` is always a `tibble`. You can see that none of the variables is a `factor`. This is the default behaviour of `readr`. If you want a variable to be coded as a `factor` then you have to transform it manually, preferably with functions from the package `forcats`.

Let's have a brief look at the data set.

```{r}
summary(emissions)
```

For character variables, `summary()` does not count the frequency of different values. However, we can get this information with the function `unique()`.

```{r}
length(unique(emissions$geo))
```

The data set contains measurements for 33 EU countries. We can also ask which different types of vehicle are recorded.

```{r}
unique(emissions$vehicle)
```

## Visualization with the library `ggplot2`

The library `ggplot2` is a powerful package for data visualization. The name comes from **the grammar of graphics** and hints at a systematic approach to visualization. In a nutshell, `ggplot2` defines a statistical graphic as follows:

> A statistical graphic is a `mapping` of variables in a `data` set to `aes`thetic attributes of `geom`etric objects.

In `ggplot2`, a graphic is built up step by step, starting with a call to the core function `ggplot()`. We have to specify the following elements:

-   `data`: the data set containing the variables to be visualized.

-   `aes`: (aesthetic) attributes for the geometric object to be visualized. This can be the x and y variables, colour, shape, grouping variable etc.

-   `geom`: the geometric object we want to plot, i.e. lines, points, bars, boxes etc.

### Line plot

We start with the line plot that is particularly suited for time series. Because plotting 33 countries in one graph is too much, we first filter for France and emissions from cars.

```{r}
emissions_france <- emissions %>% 
  filter(geo == 'France' & vehicle == 'Fuel combustion in cars')
```

```{r, out.width = "90%"}
ggplot(data = emissions_france, mapping = aes(x = time, y = values))
```

This call to `ggplot()` prepares the plotting area as requested, but does not show anything because we did not specify any geometric object. All geometric objects begin with `geom_`. Every further step in building up the plot is appended to the core call by a `+`.

```{r, out.width = "90%"}
ggplot(data = emissions_france, mapping = aes(x = time, y = values)) +
  geom_line()
```

This call can be verbalized like following:

-   Take the data set `emissions` and map the following attributes:

    -   on the $x$-axis, the variable `time`
    -   on the $y$-axis, the variable `values`

-   Plot the data as a line (`geom_line()`)

In order for the plot to be useful, we should label the axes correctly (and give it a title, if no figure caption is shown). This is done by adding the function `labs()`.

```{r, out.width = "90%"}
ggplot(data = emissions_france, mapping = aes(x = time, y = values)) + 
  geom_line() +
  labs(x = 'Time', y = 'Emissions (Mio tons)', title = 'Emissions in France')
```

### Point plot

We can add points to this plot with `geom_point()`. Normally, I wouldn't do it for this time series, but I want to show this geom `r emo::ji('smile')`.

```{r, out.width = "90%"}
ggplot(data = emissions_france, mapping = aes(x = time, y = values)) + 
  geom_line() +
  geom_point() +
  labs(x = 'Time', y = 'Emissions (Mio tons)', title = 'Emissions in France')
```

If we select two countries, then a further aesthetic will be required to distinguish the time series. Let's select France and Italy.

```{r}
emissions_france_italy <- emissions %>% 
  filter(geo %in% c('France', 'Italy') & vehicle == 'Fuel combustion in cars')
```

We plot the countries using different colours. Note that we do not (yet) select the colours manually, but specify the variable that will be used to distinguish the time series. The colours are chosen one by country automatically.

```{r, out.width = "90%"}
ggplot(data = emissions_france_italy, mapping = aes(x = time, y = values, colour = geo)) + 
  geom_line() +
  geom_point() +
  labs(x = 'Time', y = 'Emissions (Mio tons)', title = 'Emissions in France and Italy', colour = 'Country')
```

The legend comes for free! We can change the title of the legend by setting `colour = 'Country'` in the call to `labs()`.

## Histogram

Let's have a look at the distribution of emissions in the year 2018. We have to filter the data first.

```{r}
emissions_2018 <- emissions %>% 
  filter(time == '2018-01-01')
```

We plot the data in a histogram that shows the absolute frequencies of the data (i.e. how many data points fall in a particular interval of emissions). It shows the distribution of a continuous variable. For a histogram, we only specify the x variable, the frequencies are calculated by `geom_histogram()` directly. We specify 25 bins (intervals). If you are not familiar with this kind of statistical summaries, please have a look at the Appendix A in @ModernDive and read the part A.1.5 Distribution.

```{r}
ggplot(data = emissions_2018, mapping = aes(x = values)) +
  geom_histogram(bins = 25)
```

## Boxplot

A boxplot calculates some prominent statistics of a data set and plots them in the form of a box with 'whiskers' (thus also called box-and-whiskers plot). Basically, it is the same as calculating the `summary()` (five-numbers: min, max, 25%, 50% and 75% quantiles), but as a figure. If you are not familiar with this kind of statistical summaries, please have a look at the Appendix A in @ModernDive and read the part A.1.4 Five-number summary.

Let's have a look at this kind of summary plot. How are the emissions distributed by year? We have to convert `time` to a factor variable to display the data correctly (try out what happens if you don't convert it).

```{r, out.width = "90%"}
ggplot(data = emissions, mapping = aes(x = factor(time), y = values)) +
  geom_boxplot()
```

Hmmm, the labels on the $x$-axis are ugly. Let's tune them a little (we will do more tuning in later sessions).

```{r, out.width = "90%",}
ggplot(data = emissions, mapping = aes(x = factor(time), y = values)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90))
```

### Barplot

The last geom we want to see is `geom_bar()`. We would like to know how many data entries does `emissions` contain per vehicle.

```{r, out.width = "90%"}
ggplot(data = emissions, mapping = aes(x = vehicle)) +
  geom_bar()
```

Admittedly, a boring plot `r emo::ji('smile')`, the number of entries is identical.

## Practice on your own!

<ol class ="exercises">
1. The histogram, the boxplot and the barplot plotted above are not labelled correctly. Correct the axis labels and find good titles for the graphs.

2. Plot the time series of GDP from the data set `gapminder` for France and Germany. Filter the data like this:
<br>
`france_germany <- gapminder %>% filter(country %in% c('France', 'Germany'))`

3. Plot the life expectancy vs. GDP in 2007, use the data set `gapminder`. Pick the code for filtering the data in task \@ref(firstplot). Use the aesthetics colour and size. Have an educated guess on how to change the title of the legends (or google for it `r emo::ji('smile')`).

4. How is the GDP distributed in Africa and Europe in 2007? Use the data set `gapminder`. Filter the data like this:
<br>
`africa_europe <- gapminder2007 %>% filter(continent %in% c('Africa', 'Europe'))`.
<br>
Plot the data as a histogram and use the aesthetic fill instead of colour to distinguish between continents.

5. How is the GDP distributed on different continents in 2007? Use the data set `gapminder`. Plot the data as a boxplot.

6. How many data points does `gapminder` contain per continent? Visualize in a barplot.
</ol>

## Reading assignment

Chapter 2.1 in @ModernDive

## Turning in your work
- Save your R Notebook as an *.Rmd file. 
- Upload your R Notebook to ILIAS. You don't need to upload the .nb.html file. You will find an upload option in today's session.
- You should receive a solution file after your upload. **Be sure to upload before the deadline!**



## Additional reading

Chapters 2.2 to 2.9 in @ModernDive
