# Explorative workflow with `tidyverse` {#tidyverse}

```{block, type='rmdoutcomes'}
- Name core packages in `tidyverse`
- Apply a simple explorative workflow (read, summarize, plot) with `tidyverse`
- Use functions from `dplyr` for data wrangling
```

`tidyverse` is a collection of R packages for data analysis (https://www.tidyverse.org/). It shares a common philosophy about how data should be structure and grammar of data manipulation and visualisation. Although it might sound like something alien, `tidyverse` is a regular part of R and its functions can be mixed with base R functions.

The best introduction to `tidyverse` is **r4ds**: "R for Data Science" [@r4ds]. You can read it for free here (https://r4ds.had.co.nz/).

## Core packages
`tidyverse` comprises 8 core packages that are installed when you call `install.packages('tidyverse')`:

Packages | Description|
:-------- | :------------ |
`ggplot2`| data visualization |
`dplyr`| data transformation |
`tidyr`| data cleaning |
`readr`| importing data |
`purrr`| functional programming|
`tibble`| extension of `data.frame` |
`stringr`| functions for strings, i.e. text variables  |
`forcats`| functions for `factor` |

Every packages has a Cheat Sheet, an overview of its functions. To get a package's cheat sheet click on its name (https://www.tidyverse.org/packages/), scroll down to the section Cheatsheet.

Besides its core packages, `tidyverse` also installes a long list of supplementary packages that you can find here: https://www.tidyverse.org/packages/


## Exploratory data analysis

Exploratory data analysis is a very important first step in data analysis. Before using any advanced statistical method, exploratory analysis is a must have. It comprises roughly the following steps:

- import and inspect data
- clean (tidy) data if necessary
- summarize it and create new variables if necessary
- plot as many different plots as possible to get a good overview about patterns and data distribution

### Read data, revisited

We load the library `tidyverse` first.
```{r, message = F, warning = F}
library(tidyverse)
```

Last time we used the function `read_delim()` to import data into R. This function is the most general from a whole family of functions, all starting with `read_*`: `read_csv()`, `read_csv2()` etc. They all have their own parameters that you need to verify on the respective help pages if you want to use them.

For this exploratory data analysis we will use data from the German Meteo Service (Deutscher Wetterdienst) that I downloaded on 2020-05-24 (https://www.dwd.de/DE/leistungen/klimadatendeutschland/klimadatendeutschland.html). The data set contains hourly measurements of the relative air humidity (%) and air temperature (°C) for three meteo stations, namely Hof, Frankfurt and Köln-Bonn. The data is named `r file.name("meteo.csv")`.

```{r}
temp_humid <- read_delim('data/meteo.csv', delim = ';',    trim_ws = T)
```

`read_delim()` reports on reading the data about the variables it recognizes. This is a good hint to spot for possible problems. Are numerical variables read as `<dbl>`? Are characters recognized as `<char>` etc. In the code above, the parameter `trim_ws = T` removes leading zeroes.

Let's have a showr glimps of the data.
```{r}
temp_humid
```
The data set contains the following variables:

Variable | Description |
:--------|:---------------|
STATIONS_ID | ID of the meteo station |
MESS_DATUM | date and time of the measurement, formatted as  yyyymmddhh |
QN_9 | quality flag |
TT_TU | air temperature in 2 m height in °C |
RF_TU | relative air humidity in % |
eor | end of record (i.e. end of line)

`read_*` always returns a `tibble`.
```{r}
class(temp_humid)
```

### Date and time made easy
A very useful package to handle date and time is called `lubridate`. It is not part of core packages in `tidyverse` but is installed with the long list of additional packages. We will use it to convert the variable `MESS_DATUM` to a real date-time variable.

The function `ymd_h()` converts character vectors to date-time objects provided they have the format **y**ear, **m**onth, **d**ay, **h**our. There are other function for different other formats; consult help.

```{r, message = F, warning = F}
library(lubridate)

temp_humid$MESS_DATUM <- ymd_h(temp_humid$MESS_DATUM)

temp_humid
```

After conversion, the variables is recognised as `<dttm>` for date-time.

### Summarize
The three meteo station have the following IDs:
```{r}
station_ids <-  c('2261' = 'Hof', '1420' = 'Frankfurt', '2667' = 'Koeln')
```

We want to know how many measurements per station the data set contains.

```{r}
temp_humid %>% 
  group_by(STATIONS_ID) %>% 
  count()
```

The operator `%>%` is called **pipe** and is pronounced as **and then**. The code `temp_humid %>% group_by(STATIONS_ID) %>% count()` can be read as: take the object `temp_humid`, group it by `STATIONS_ID` and count the measurments in each group. The pipe operator comes from the package `magrittr` (https://magrittr.tidyverse.org/). It is a core operator in `tidyverse` and makes the code more readable and easier to follow for humans. Perhaps not in the beginning, but very soon `r emo::ji('nerd_face')`.

### The grammar of data manipulation -- `dplyr`
The function `count()` is part of the library `dplyr`, a collection of functions all named after verbs. Thus, it is easy to imagine what the function does `r emo::ji('smile')`). The 5 core functions are:

Function | Meaning |
:-----------|:-----------|
`filter()` | filter data according to their values  |
`arrange()` | arrange rows |
`select()` | select variables according to their names |
`mutate()` | create new variables, possibly using other variables  |
`summarize()` | summarize data with different functions |

If we want to know how many measurements were recorded for a particular meteo station, we first filter for its ID:

```{r}
temp_humid %>% 
  filter(STATIONS_ID == '2667') %>%
  count()
```

The function `filter()` accepts logical tests. For every row in `STATION_ID`, `==` checks whether the entry equals 2667. `==` is a logical operator and means *is the left side equals the right sight*.  If this is the case, then `==` returns `TRUE` otherwise it returns `FALSE`. `filter()` selects only those rows where `TRUE` was returned. Other useful logical operators are:

Operator | Meaning |
:-----------|:-----------|
`>` | is the left side larger than the right side?|
`>=` | is the left side larger or equal the right side? |
`!=`| are left and right sides unequal? |

More logical and boolean operators are handeled in the tutorials (see below) and on help pages of `filter()`.

We can combine several test with the operator **or** `|`, for example. Here, we want to filter all rows containing either ID 2667 or ID 2261:
```{r}
temp_humid %>% 
  filter(STATIONS_ID == '2667' | STATIONS_ID == '2261') %>%
  group_by(STATIONS_ID) %>% 
  count()
```

The same can be achieved by excluding the third station:
```{r}
temp_humid %>% 
  filter(STATIONS_ID != '1420') %>%
  group_by(STATIONS_ID) %>% 
  count()
```

As an alternative, we can use the operator `%in%` which checks whether the row contains one of the entries in a vector.
```{r}
temp_humid %>% 
  filter(STATIONS_ID %in% c('2667', '2261')) %>%
  group_by(STATIONS_ID) %>% 
  count()
```

### Visualise

We plot the time series and use a trick to split them along three different plots with the function `facet_wrap()`. It needs a variable to separate the data into plots and we chose `STATIONS_ID`. The splitting variable must be preceeded by `~`.

```{r, out.width = "90%"}
ggplot(data = temp_humid, aes(x = MESS_DATUM, y = TT_TU)) + 
  geom_line() +
  facet_wrap(~STATIONS_ID, nrow = 3) +
  labs(x = 'Time', y = 'Temperature (°C)')
```

### New variables with `mutate()`
We want to calculate the monthly means and standard deviations of the air temperature and humidity. First, we need to generate the temporal information, namely year and month that will be used to group the temperature values to calculate `mean()` and `sd()`. This can be achieved with the functions `year()`and `month()` from library `lubridate`. The function `mutate()` can create new variables in a data object. 

```{r}
temp_humid <- temp_humid %>% 
  mutate(year = year(MESS_DATUM),
         month = month(MESS_DATUM))

temp_humid
```

In the next step, we create a new data set and calculate the means and standard deviations. In order to get them *by station, year and month*, we group the data accordingly. To group by several variables, just enumerate them with a comma (no quotation or `c()` necessary).

```{r}
monthly_means <- temp_humid %>%
  group_by(STATIONS_ID, year, month) %>% 
  summarize(mean_T = mean(TT_TU), mean_RH = mean(RF_TU),
            sd_T = sd(TT_TU), sd_RH = sd(RF_TU))

monthly_means
```

The new object `monthly_means` is a grouped `tibble`, indicated by *grouped_df* in the output of `str()` that shows the structure of an object.

```{r}
str(monthly_means)
```

Some calculations are better done on ungrouped data. Therefore, we remove the grouping. This does not change the data itself.

```{r}
monthly_means <- ungroup(monthly_means)
```

To plot the monthly data, we need a proper monthly date object. We will attribute the monthly means to the first of the respective month. Again, `lubridate` helps with this task. The function `parse_dat_time()` is a genral function taking a character string and returning a date-time object. We need to "glue" the variables `year` and `month` together with `paste0()` (yes, it is a zero, not an O!) to form such a string and specify that `orders = 'ym'`, i.e. year before month. Finally, we `relocate()` the new variable `year_month` before the variable `year` for convenience (if not, it will be created as the last varialbe in the data set).

```{r}
monthly_means <- monthly_means %>%
  mutate(year_month = parse_date_time(paste0(year, month), orders = 'ym', tz = 'CET')) %>% 
  relocate(year_month, .before = year)

monthly_means
```
Now, we can plot the mean air temperature.
```{r, out.width = "90%"}
ggplot(data = monthly_means, aes(x = year_month, y = mean_T, col = factor(STATIONS_ID))) + 
  geom_line() + 
  labs(x = 'Time', y = 'Temperature (°C)', color = 'Meteo station')
```

We can also visualise the standard deviations.
```{r, out.width = "90%"}
ggplot(monthly_means, aes(x = year_month, y = mean_T, ymin = mean_T - sd_T, ymax = mean_T + sd_T)) +
  geom_errorbar() +
  geom_point() +
  facet_wrap(~STATIONS_ID, nrow = 3) + 
  labs(x = 'Time', y = 'Temperature (°C)')
```

Or use a semi-transparent band to show the variability (standard deviation).

```{r, out.width = "90%"}
ggplot(monthly_means, aes(x = year_month, y = mean_T, ymin = mean_T - sd_T, ymax = mean_T + sd_T)) +
  geom_ribbon(alpha = 0.5) +
  geom_line() +
  facet_wrap(~STATIONS_ID, nrow = 3) + 
  labs(x = 'Time', y = 'Temperature (°C)')
```

One last detail. The titles on top of the facets show the station IDs. When you are not an employee of German Meteo Service, you probably do not know them by hart. It is better to use the city names. The vector `station_ids` is a so called named vector and has the right structure to change the titles in the facets: it assignes to every id its city name, i.e. 2261 = 'Hof'.

```{r}
station_ids
```

We use `station_ids` to change the titles:

```{r, out.width = "90%"}
ggplot(monthly_means, aes(x = year_month, y = mean_T, ymin = mean_T - sd_T, ymax = mean_T + sd_T)) +
  geom_ribbon(alpha = 0.5) +
  geom_line() +
  facet_wrap(~STATIONS_ID, nrow = 3, labeller = labeller(STATIONS_ID = station_ids)) + 
  labs(x = 'Time', y = 'Temperature (°C)')
```

## Practice on your own!

1. Plot the means and standard deviations of the air humidity instead of air temperature.

1. Do the tutorials "Work with data" from the Primers collection by RStudio Cloud. You can access the tutorials here: https://rstudio.cloud/learn/primers/2

1. Do the tutorials "Visualize Data" from the Primers collection by RStudio Cloud. You can access the tutorials here:https://rstudio.cloud/learn/primers/3


## Reading assignment
Read chapter 3 (to 3.5) in @ModernDive


<!-- ## Turning in your work -->
<!-- - Save your R Notebook and download the .Rmd file to your computer. You don't need to download the .nb.html file. -->
<!-- - Upload your R Notebook to ILIAS. You will find an upload option in today's session. -->
<!-- - You should receive a solution file after the submission. -->

<!-- ```{block, type='rmdalert'} -->
<!-- Be sure to upload before the deadline! -->
<!-- ``` -->

## Additional reading and videos

- More information on statistical graphical summaries and geoms: R4DS @r4ds: Chapter 5 "Data transformation"

- A live exploratory data analysis by the main author of `tidyverse`, Hadley Wickham. Really informative, but he types too fast `r emo::ji('smile')`.

<iframe width="639" height="302" src="https://www.youtube.com/embed/go5Au01Jrvs" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>
