# Additional exercises

## Introduction to R and RStudio Server Pro

### Rob's account book {#rob1}

The young master student Rob Stat thinks seriously about his mother's advice to monitor his expenses. He begins by writing down what he spends during the week in the Mensa:

<br>

```{r, results='asis', echo = F}

tab <- tibble(`Day of week` = c('Monday', 'Tuesday', 'Wednesday', 'Thrusday', 'Friday'), `Amount spent (€)` = c(2.57, 2.90, 2.73, 3.23, 3.90))
knitr::kable(tab, 
             #format = 'latex',
             caption = "Rob's account book", booktabs = T,
             format.args = list(decimal.mark = '.')) %>%
  kable_styling(bootstrap_options = "striped", full_width = F,
                position = "left")
```


- Generate a vector of Rob's expenses and assign it to the variable `expenses`. Use the function `c()` and use the numeric expenses only, not the days of the week.
- How much did Rob spend during this week? Use the function `sum()`.
- Rob seems to have spent the smallest amount on Tuesday. How much would he have spent if he paid that much every day of the week? Use the array notation with the square brackets.

Unfortunately, Rob misspelled the amount on Tuesday. Actually, he invited his girl friend for lunch and paid 7.95 € instead of 2.90 €.

- Correct Rob's typo.
- How does the result change?

### Missing values {#rob2}

R codes missing values as `NA`. Rob ate in the Mensa last Monday, but forgot to write down the amount.

```{r, results='asis', echo = F}

tab <- tibble(`Day of week` = c('Monday, 9 March', 'Tuesday, 10 March', 'Wednesday, 11 March', 'Thrusday, 12 March', 'Friday, 13 March', 'Monday, 16 March'), `Amount spent (€)` = c(2.57, 2.90, 2.73, 3.23, 3.90, NA))
knitr::kable(tab, 
             #format = 'latex',
             caption = "Rob's account book, cont.", booktabs = T,
             format.args = list(decimal.mark = '.')) %>%
  kable_styling(bootstrap_options = "striped", full_width = F,
                position = "left")
```

- How does NA change the calculated sum?
- Read what happens when the data contains NAs by calling help on `sum`, i.e. type `?sum` in R console.
- Correct your call to `sum()` accordingly.

### Your very first plot {#firstplot}
In particular in the beginning of learning R you should not forget why you are doing it. Because R is really beautiful and you want to analyse and learn from real data. 

Even if you don't fully understand the following code, just copy and paste it into your .R file and let it run!

```{r, echo = T, results = 'hide', fig.show='hide'}
library(tidyverse)
library(gapminder)

gapminder2007 <- gapminder %>% 
  filter(year == 2007)

ggplot(gapminder2007, aes(x = gdpPercap, y = lifeExp, color = continent, size = pop)) +
  geom_point() +
  scale_x_log10() +
  xlab('GDP per capita') +
  ylab('Life expectancy') +
  labs(title = 'Gapminder data for the year 2007')
```

- What is the data set about. Use help like this `?gapminder`.
- What do the colours represent?
- What does the size of the circles represent?
- How would you describe the relationship between the GDP per capita and Life expectancy?

## The big practical: importing, wrangling, summerizing and plotting 

### Temperature along the Dutch coast
The file `r file.name("Temperatur.csv")` from the book by @Zuur2009a contains measurements of temperature, salinity and content of chlorophyll a at 31 locations along the Dutch coast. You can download the data set [here](https://highstat.com/index.php/a-beginner-s-guide-to-r). The data is provided by the Dutch institute RIKZ (monitoring program MWTL: Monitoring Waterstaatkundige Toestand des Lands), was measured between 1990 and 2005 between 0 and 4 times per month depending on the season. 

- Read the file `r file.name("Temperatur.csv")` into R.
- Convert the column `Date` to a proper date format. Use the library `lubridate`.
- Calculate the number of measurements, mean and standard deviations of temperature per monitoring station. Hint: use `n()` inside `summarize()` to get the number of measurements.
- Calculate the number of measurements, mean and standard deviations of temperature per month.
- Plot the mean monthly temperature as a line and add the standard deviation as a band around it.
- Label the axis appropriately.
- Save you graph as a pdf file.


### Temperature along the Dutch coast, revisited
- Calculate the monthly means and standard deviations per monitoring station. Hint `group_by(Station, Month)`.
- Plot the means with an error band in different plots. Hint: use `facet_wrap())`.
- Save you graph as a pdf file.


### Excel data turns tidy
We will import and tidy World Development Indicators data downloaded from the [World Bank](https://databank.worldbank.org/source/world-development-indicators/preview/on#) on 2021-06-09 for 20 countries. This is an extract only and more data is available.

This exercise will show you how to load excel data directly without converting it to .csv file. The format of the data is a typical non-tidy one and you will wrangle it to a tidy `tibble`. The file is called `r file.name('Data_Extract_From_World_Development_Indicators.xlsx')`.

- First, read the help pages of the function `read_xlsx()` in the library `readxl`. Pay attention to how you can specify the name of the excel sheet you want to read into R.
- Open the excel sheet and look through the data carefully. How are NAs coded? Which data sheet do you need to read?
- Read the excel file into R. Call it `wdi`.
- This data set is not tidy. In particular, the year is coded as column name. Those column names contain the year twice, once as a number and once as [YR NUMBER]. We rename the columns first.
```{r, eval = F}
wdi <- wdi %>% 
  rename_with(.fn = function(x) str_sub(x, start = 1, end = 4), .cols = starts_with('20'))
```

What does this code mean? Read the help pages for functions `rename_with()`, `str_sub()` and `starts_with()`.

- Pivot the data set to a tidy format: variables in columns and measurements in rows. Use `pivot_longer`.
```{r, eval = F}
wdi <- wdi %>%
  pivot_longer(names_to = 'year', values_to = 'indicator_value', cols = starts_with('20')) %>% 
  mutate(year = as.numeric(year))

wdi
```

  What does this code mean? Read the help pages for functions `pivot_longer()` and `as.numeric()`.
  Why is it necessary to convert `year` with `as.numeric()`?

- Filter for the indicator `GDP (current US$)` and plot the data as time series. Hint: You can also filter for the indicator's code; look it up in the excel file. Label the axis appropriately.